---
title: "Custom Rules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Rules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Policies are lists of `shieldr_rule` objects plus thresholds. You can start
with a preset and append domain-specific rules.

```{r}
library(llmshieldr)
```

For the source model behind the built-in presets, see
`vignette("policy-design", package = "llmshieldr")`.

## Rule Fields

Every rule has the same shape:

- `id`: unique rule identifier
- `pattern`: regex pattern, or `NULL`
- `fn`: R predicate function, or `NULL`
- `owasp`: OWASP LLM category such as `llm02`
- `severity`: `low`, `medium`, `high`, or `critical`
- `action`: `allow`, `redact`, or `block`
- `description`: human-readable explanation

Exactly one of `pattern` or `fn` must be supplied. Regex rules produce match
spans that can be redacted. Function rules are useful when the condition is
easier to express in R.

## Numbers and Thresholds

Severity maps to risk score contributions:

| Severity | Contribution |
| --- | ---: |
| `low` | 0.1 |
| `medium` | 0.3 |
| `high` | 0.6 |
| `critical` | 1.0 |

Findings are summed and capped at `1.0`. A policy's thresholds then decide the
final action. Defaults are `redact_at = 0.4` and `block_at = 0.75`.

```{r}
policy <- policy_preset("enterprise_default")
policy$thresholds
```

## Regex Rules

Regex rules are the simplest way to redact or block recognizable text.

```{r}
policy <- add_rule(
  policy,
  id = "llm02.ticket_id",
  pattern = "\\bTICKET-[0-9]{6}\\b",
  owasp = "llm02",
  severity = "medium",
  action = "redact",
  description = "Internal support ticket identifier."
)

scan_prompt("Summarize TICKET-123456 for the support team.", policy)
```

## Function Rules

Function rules let you express checks that are easier to write in R than in a
single regular expression.

```{r}
contains_student_address <- function(text) {
  grepl("\\bstudent\\b", text, ignore.case = TRUE) &&
    grepl("\\bhome address\\b", text, ignore.case = TRUE)
}

education <- policy_preset("education_safe")
education <- add_rule(
  education,
  id = "llm02.student.address",
  fn = contains_student_address,
  owasp = "llm02",
  severity = "high",
  action = "redact",
  description = "Student home address reference."
)

scan_prompt("The student home address appears in the form.", education)
```

## Industry Examples

Healthcare and life sciences often add identifiers beyond generic PII.

```{r}
pharma <- policy_preset("pharma_gxp")
pharma <- add_rule(
  pharma,
  id = "llm02.site_id",
  pattern = "\\bSITE-[0-9]{3}\\b",
  owasp = "llm02",
  severity = "medium",
  action = "redact",
  description = "Clinical trial site identifier."
)
```

Finance workflows often tighten language around recommendations and promises.

```{r}
finance <- policy_preset("finance_strict")
finance <- add_rule(
  finance,
  id = "llm09.promissory_return",
  pattern = "(?i)guaranteed\\s+(alpha|profit|return)",
  owasp = "llm09",
  severity = "critical",
  action = "block",
  description = "Promissory investment performance claim."
)
```

## Rule Inventory

Use `list_rules()` to inspect a policy before deployment.

```{r}
list_rules(policy)
```

The resulting table includes `has_pattern` and `has_fn`, which make it easy to
audit whether a policy is mostly regex-based, function-based, or mixed.
