---
title: "Gemini Developer API workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gemini Developer API workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`llmshieldr` uses the suggested `ellmer` package for Gemini. Install `ellmer`
and set `GEMINI_API_KEY` or `GOOGLE_API_KEY` outside your script. Live examples
are disabled during vignette builds and CRAN checks.

The Gemini Developer API sends admitted prompt and context text to Google.
The examples use stable models currently listed for free-tier use. Google
determines eligibility, quotas, model availability, and data-use terms by
account, project, and region. Review
[current pricing](https://ai.google.dev/gemini-api/docs/pricing),
[rate limits](https://ai.google.dev/gemini-api/docs/rate-limits), current
terms, and your organization's data rules before using private content. The
current pricing table marks free-tier content as eligible for product
improvement.

## Guard a public prompt

```{r, eval = FALSE}
library(llmshieldr)

result <- secure_chat(
  "Summarize the public release note in three bullets.",
  provider = "gemini",
  model = "gemini-3.8-flash",
  reviewer_model = "gemini-3.5-flash-lite",
  checks = "both",
  show_stats = TRUE
)
result$output
result$audit$token_estimate
```

The common interface is not limited to Gemini and Ollama: `provider` accepts any
name supported by `ellmer::chat()`. The `"gemini"` spelling above is retained
as a convenience alias for ellmer's `"google_gemini"` provider name.

The assistant and reviewer are separate chats. `show_stats = TRUE` reports
elapsed time, assistant and reviewer token use where `ellmer` exposes it,
estimated scanned-text tokens otherwise, and network status. The `ellmer` API does
not expose reliable per-call wire byte counts, so upload/download rates are
shown as unavailable instead of inferred from character counts.

## Admit only approved context

```{r, eval = FALSE}
guardrails <- policy("enterprise_default")
admission <- context_policy(
  required_columns = c("document_id", "source", "tenant"),
  tenant_id = "public",
  trusted_sources = "public_kb"
)
context <- data.frame(
  text = c("Public release notes.", "Private tenant note."),
  source = c("public_kb", "private_kb"),
  document_id = c("release-1", "private-1"),
  tenant = c("public", "other"),
  stringsAsFactors = FALSE
)

result <- secure_chat(
  "Summarize the release notes.",
  provider = "gemini",
  model = "gemini-3.8-flash",
  reviewer_model = "gemini-3.5-flash-lite",
  policy = guardrails,
  context = context,
  context_policy = admission,
  checks = "both"
)
```

Apply tenant and document permissions in the retrieval query too.
`context_policy()` is a second admission check before model submission.

## Bring an existing Gemini chat

```{r, eval = FALSE}
assistant <- ellmer::chat_google_gemini(model = "gemini-3.8-flash", echo = "none")
reviewer <- ellmer::chat_google_gemini(model = "gemini-3.5-flash-lite", echo = "none")
secure_chat(
  "Explain a public technical term.",
  chat = assistant,
  reviewer = reviewer,
  checks = "both",
  show_stats = TRUE
)
```

Ollama remains available through `secure_chat(provider = "ollama")` for local
workflows. The older provider-specific wrappers are deprecated compatibility
aliases. Model availability and quotas vary by project and may change.
