---
title: "Gemini Developer API workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gemini Developer API workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`llmshieldr` uses the suggested `ellmer` package for Gemini. Install `ellmer`
and set `GEMINI_API_KEY` or `GOOGLE_API_KEY` outside your script. Live examples
are disabled during vignette builds and CRAN checks.

The Gemini Developer API sends admitted prompt and context text to Google.
Google states that free-tier content may be used to improve its products.
Review [current pricing](https://ai.google.dev/gemini-api/docs/pricing),
[rate limits](https://ai.google.dev/gemini-api/docs/rate-limits), and your
organization's data rules before using private content.

## Guard a public prompt

```{r, eval = FALSE}
library(llmshieldr)

result <- shield_gemini(
  "Summarize the public release note in three bullets.",
  model = "gemini-2.5-flash",
  reviewer_model = "gemini-2.5-flash-lite",
  checks = "both",
  show_stats = TRUE
)
result$output
result$audit$token_estimate
```

The assistant and reviewer are separate chats. `show_stats = TRUE` reports
elapsed time, token use or estimates, and network status. The `ellmer` API does
not expose reliable per-call wire byte counts, so upload/download rates are
shown as unavailable instead of inferred from character counts.

## Admit only approved context

```{r, eval = FALSE}
guardrails <- policy(
  "enterprise_default",
  overrides = list(trusted_sources = "public_kb")
)
context <- data.frame(
  text = c("Public release notes.", "Private tenant note."),
  source = c("public_kb", "private_kb"),
  tenant = c("public", "other"),
  stringsAsFactors = FALSE
)

result <- shield_gemini(
  "Summarize the release notes.",
  policy = guardrails,
  context = context,
  context_authorize = function(row) identical(row$tenant[[1]], "public"),
  checks = "both"
)
```

Apply tenant and document permissions in the retrieval query too. The row
authorization callback is a second check before model submission.

## Bring an existing Gemini chat

```{r, eval = FALSE}
assistant <- ellmer::chat_google_gemini(model = "gemini-2.5-flash", echo = "none")
reviewer <- ellmer::chat_google_gemini(model = "gemini-2.5-flash-lite", echo = "none")
secure_chat(
  "Explain a public technical term.",
  chat = assistant,
  reviewer = reviewer,
  checks = "both",
  show_stats = TRUE
)
```

Ollama remains available through `shield_ollama()` for local workflows. Model
availability and quotas vary by project and may change.
