---
title: "Operations and Audit Logs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Operations and Audit Logs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE)
```

```{css, echo = FALSE, eval = TRUE}
.llmshieldr-info-box {
  border-left: 4px solid #2f80ed;
  background: #f3f8ff;
  padding: 1rem 1.15rem;
  margin: 1.5rem 0;
  border-radius: 0.35rem;
}

.llmshieldr-info-box h2,
.llmshieldr-info-box h3,
.llmshieldr-info-box h4 {
  margin-top: 0;
}

.llmshieldr-info-box p:last-child,
.llmshieldr-info-box ul:last-child,
.llmshieldr-info-box ol:last-child {
  margin-bottom: 0;
}
```

This article collects the operational guidance that matters once `llmshieldr`
is part of an application. It is not a certification checklist and does not
make a workflow production-ready by itself.

## Before Deployment

- Define the application threat model.
- Identify sensitive data classes: PII, PHI, credentials, financial data,
  customer records, intellectual property, and regulated text.
- Choose a built-in policy as a starting point.
- Add organization-specific rules and negative tests.
- Decide whether semantic review is required and which reviewer model or
  function is allowed.
- Keep the default reviewer-error action (`"block"`) or explicitly select
  `policy_controls(on_reviewer_error = "rules_only")` after assessing the risk.
- Decide `policy_controls()` behavior for prompt blocks, context blocks,
  output blocks, refusals, and human escalation.
- Configure `scanner_options()` for token limits, URL host policy, topic bans,
  and language policy when those controls matter.
- Choose a redaction strategy for logs and user-visible text.
- Evaluate detection and false-positive rates on representative data.
- Review audit log contents and storage controls.
- Confirm downstream systems still escape, validate, and authorize inputs.

## Runtime Controls

- Scan user prompts before model calls.
- Scan retrieved context before prompt assembly.
- Drop, refuse, or escalate blocked context rows according to policy.
- Supply trustworthy source labels and, where needed, a row-level authorization
  function. Admitted row identifiers and source labels appear in prompt
  assembly and audit metadata.
- Scan model output before display, storage, or downstream tool use.
- Validate tool-call names and arguments before execution.
- Scan tool outputs before they re-enter model context.
- Scan the complete buffered stream before releasing its cleaned text. Do not
  display unscanned chunks to users.
- Apply rate guards appropriate to the environment.
- Use strict rate-guard reservation for shared or bursty workloads.
- Record metadata-only audit logs to access-controlled storage.
- Monitor warnings, blocked requests, reviewer parse errors, and escalation
  counts.

::: {.llmshieldr-info-box}
## Audit Log Sensitivity

`secure_chat()` creates metadata-only audits by default. Report text, finding
excerpts, reviewer error details, and the cleaned prompt and raw output are
removed. Source labels, row indices, rule IDs, positions, token counts, and
other metadata may still identify sensitive activity. Treat audit output as
sensitive application telemetry.

To retain content in memory, call `secure_chat(..., audit_content = "full")`.
Writing it also requires `write_audit_log(..., include_content = TRUE)`. Both
choices should be limited to approved debugging and protected storage.

Supported formats:

- JSON Lines: append-only operational logs with nested audit structure.
- CSV: one row per finding for spreadsheets and simple dashboards.
- RDS: exact R object preservation for local debugging.
:::

## Example JSONL Shape

This abbreviated example shows the default metadata-only form. Actual records
also contain timestamps and other report metadata.

```json
{
  "input_report": {
    "action": "redact",
    "risk_score": 0.3,
    "policy": "enterprise_default",
    "checks": "rules",
    "findings": [
      {
        "rule_id": "llm02.pii.email",
        "owasp": "llm02",
        "owasp_edition": "LLM02:2026",
        "taxonomy_version": "OWASP-LLM-Top-10-2026",
        "severity": "medium",
        "action": "redact",
        "source": "rules"
      }
    ],
    "metadata": {
      "stage": "prompt",
      "taxonomy_version": "OWASP-LLM-Top-10-2026"
    }
  },
  "output_report": null,
  "context_reports": null,
  "tool_reports": null,
  "prompt_clean": null,
  "output_raw": null,
  "elapsed_ms": 12,
  "token_estimate": 8,
  "action": "redact",
  "content_mode": "metadata"
}
```

## Example CSV Shape

| stage | context_row_index | context_source | tool_name | conversation_role | reviewer_error_count | report_index | action | risk_score | rule_id | owasp | owasp_edition | severity | source |
| --- | ---: | --- | --- | --- | ---: | ---: | --- | ---: | --- | --- | --- | --- | --- | --- |
| input | NA | NA | NA | NA | 0 | 1 | redact | 0.3 | llm02.pii.email | llm02 | LLM02:2026 | medium | rules |
| context | 2 | approved_docs | NA | NA | 0 | 2 | block | 1.0 | llm01.injection.indirect | llm01 | LLM01:2026 | critical | rules |
| tool | NA | NA | search_docs | NA | 1 | 1 | redact | 0.3 | llm02.pii.email | llm02 | LLM02:2026 | medium | scanner |

The `owasp` field keeps the stable lowercase category code; the accompanying
`taxonomy_version` identifies the 2026 mapping.

## Execution Statistics

Every exported function accepts `show_stats = TRUE` and emits an execution
message. Local constructors and rule-based scans report elapsed time and no
network use; token counts appear when meaningful. Guarded chat calls report
assistant provider token counts when `ellmer` exposes them, otherwise an
estimate. Semantic reviewer provider tokens appear on a separate line when
available.
Reviewer calls may add network use. For arbitrary chat functions, the package
cannot always determine whether a network request occurred and reports
`unknown`.

```{r, eval = FALSE}
scan_prompt("Summarize this public note.", show_stats = TRUE)
result <- shield_gemini("Summarize this public note.", show_stats = TRUE)
```

Upload and download rates are shown only when the package can measure bytes
and request time. `remote_reviewer(show_stats = TRUE)` reports estimated JSON
request-body bytes and response-body bytes over the HTTP request duration;
these omit headers, TLS, and other wire overhead. `ellmer` provider transports
do not currently expose transfer-byte counts to `llmshieldr`, so their rates
appear as `unavailable`. Statistics are messages and do not alter return values.

## Storage Checklist

- Store audit logs outside public web roots.
- Apply filesystem or object-storage access controls.
- Encrypt logs at rest when they may contain sensitive data.
- Define retention and deletion rules.
- Avoid sending raw audit logs to third-party observability tools unless they
  are approved for the data class.
- Keep the metadata-only default for long-term retention and evaluate whether
  source labels, row indices, or finding positions should be retained.
- Include context row indexes and source identifiers when investigating RAG
  incidents, but avoid storing full source documents unless required.
- Review `reviewer_error_count` and `reviewer_error_types` when semantic
  checks are enabled; malformed reviewer JSON is a safety signal.
- Treat hash-redacted findings as sensitive metadata because deterministic
  hashes can still link repeated values across logs.

## Operational Controls

- Pin package versions.
- Pin reviewer model versions where possible.
- Keep policies under version control.
- Review changes to built-in or custom rules.
- Add regression tests for every production rule.
- Review false positives and false negatives regularly.
- Keep a manual override or human escalation path for high-impact workflows.

## Do Not Rely On `llmshieldr` Alone For

- Legal, medical, financial, or compliance decisions.
- High-impact automated actions.
- Tool calls that modify external systems.
- Complete PII/PHI discovery.
- Malware detection or sandboxing.
- Full jailbreak resistance.
- Distributed rate limiting.
- Provider or infrastructure trust.
