---
title: "Policy Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Policy Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette explains how `llmshieldr` policies are assembled, what sources
the built-in policies draw from, and how the numeric scores become actions.

```{r}
library(llmshieldr)
```

## Design Goal

An LLM safety policy should be easy to inspect, easy to test, and easy to
extend. `llmshieldr` therefore uses explicit rule objects rather than a hidden
classifier as its default layer.

A policy is a list with these fields:

```text
shieldr_policy
    name             policy identifier stored in reports
    rules            list of shieldr_rule objects
    thresholds       redact_at and block_at numeric cutoffs
    rate_guard       optional shieldr_rate_guard environment
    trusted_sources  optional allowlist used by scan_context()
    controls         secure_chat() block/refuse/escalate/drop behavior
```

Each rule is similarly explicit:

```text
shieldr_rule
    id             stable rule identifier
    pattern        regex pattern, or NULL
    fn             R predicate function, or NULL
    owasp          OWASP LLM category
    severity       low, medium, high, or critical
    action         allow, redact, or block
    description    human-readable explanation
```

Exactly one of `pattern` or `fn` must be supplied. This keeps each rule
deterministic and makes redaction spans possible for regex rules.

## Source Model

The built-in rules are sourced from a small number of security and governance
concerns that recur across LLM applications:

- OWASP GenAI / LLM Top 10:
  <https://genai.owasp.org/llm-top-10/>
- Prompt-injection patterns: direct override language, hidden instructions,
  role confusion, and system-prompt extraction attempts
- NLP intent signals: token and stem patterns for override, secret exposure,
  harmful intent, and unusually dense directive language. Trigger seed groups
  are expanded with stems at runtime.
- Sensitive information patterns: email, phone, SSN, account numbers, medical
  record identifiers, subject IDs, API keys, bearer tokens, AWS keys, and
  credential-bearing connection strings
- Agentic risk patterns: model claims that it sent, deleted, granted,
  executed, notified, traded, or otherwise acted outside the chat boundary
- Output handling patterns: unsafe shell or SQL snippets, code execution
  markers, system-prompt structure, and high-confidence medical or financial
  claims
- RAG-specific signals: untrusted source metadata, anomalous chunk length, and
  high density of instruction words inside retrieved text
- Resource-exhaustion controls: request and token windows managed by
  `rate_guard()`, with optional strict pre-call reservation and local
  file-locking for shared guards
- Optional scanner controls: invisible Unicode format characters, encoded
  payloads, URL host policies, token limits, simple language allowlists, and
  topic bans
- Runtime surfaces: conversation scanning, tool-call scanning, tool-output
  scanning, and streaming output scanning with rolling context

These sources are intentionally conservative. They are meant to catch common
failure modes in R workflows before prompts, retrieved context, or model output
cross a trust boundary.

## Built-In Policy Construction

Every built-in policy is assembled in `policy()`.

```text
enterprise_default
    prompt injection rules
    PII and secret rules
    system prompt extraction rules
    excessive agency rules

pharma_gxp
    enterprise_default
    MRN and clinical subject identifiers
    diagnosis and treatment claim language
    code-safety checks
    stricter thresholds: redact_at = 0.3, block_at = 0.6

finance_strict
    enterprise_default
    account number rules
    financial advice and guaranteed-return language
    autonomous investment-action language
    rate_guard(max_tokens = 100000)

education_safe
    enterprise_default
    minor-related PII patterns
    academic-integrity bypass language

open_research
    injection rules
    secret rules
    higher thresholds: redact_at = 0.8, block_at = 0.95

comprehensive
    enterprise_default
    pharma_gxp additions
    finance_strict additions
    education_safe additions
    code-safety checks
    rate_guard(max_tokens = 100000)
    moderate thresholds: redact_at = 0.4, block_at = 0.7

custom
    no rules
    default thresholds

baseline
    alias for enterprise_default
```

The policy names reflect intended operating posture, not legal compliance.
For example, `pharma_gxp` adds useful GxP-style controls, but it does not by
itself make a workflow compliant with any regulation.

The `comprehensive` policy is broad rather than maximally strict. Use explicit
threshold overrides when you want pharma-tier thresholds:

```{r}
policy(
  "comprehensive",
  overrides = list(thresholds = list(redact_at = 0.3, block_at = 0.6))
)
```

## Check Modes

Scanners can run different layers depending on the workflow:

- `checks = "rules"` runs the policy's deterministic rules. Built-in policies
  include regex rules and the NLP intent rule.
- `checks = "nlp"` runs only NLP intent rules. This is the local token/stem
  path that uses `tokenizers` and `SnowballC` when installed.
- `checks = "llm"` runs only the supplied reviewer, such as `ollama_reviewer()`
  or your own reviewer function.
- `checks = "both"` runs policy rules and the supplied reviewer.

The semantic reviewer prompt is inspectable with `reviewer_prompt()`. If you
need different reviewer instructions, wrap the reviewer function or chat object
and prepend your own context before delegating to the model.

Reviewer output may be either an array of finding objects or an object with a
`findings` array. Each finding can include:

```text
rule_id              stable reviewer rule identifier
owasp                OWASP category such as llm01
severity             low, medium, high, or critical
description          human-readable explanation
confidence           optional number from 0 to 1
evidence             optional short evidence string
recommended_action   optional allow, redact, or block
span                 optional start/end offsets for redaction
```

Malformed JSON and schema issues are soft failures. The scanner warns, keeps
deterministic findings, and stores structured reviewer errors in
`report$metadata$reviewer_errors`.

## Scoring

Each finding contributes a numeric severity weight.

| Severity | Contribution |
| --- | ---: |
| `low` | 0.1 |
| `medium` | 0.3 |
| `high` | 0.6 |
| `critical` | 1.0 |

The scanner sums contributions and caps the total at `1.0`. Findings are
deduplicated before scoring. Overlapping span findings from the same source,
OWASP category, and action count as the strongest single piece of evidence
instead of stacking together. Synthetic context findings are scored separately
and capped, so source or anomaly signals cannot by themselves dominate a
report.

```text
findings:
    medium email finding      0.3
    high secret finding       0.6

risk_score = min(0.3 + 0.6, 1.0) = 0.9
```

The score is deliberately coarse. It is not a probability. It is a deterministic
severity index that helps a policy decide whether to allow, redact, or block.

## Action Resolution

Thresholds define how much risk the policy tolerates.

```text
default thresholds:
    redact_at = 0.40
    block_at  = 0.75
```

Action resolution is conservative:

```text
if any finding is critical:
    block
else if any finding's rule action is block:
    block
else if risk_score > block_at:
    block
else if any finding's rule action is redact:
    redact
else if risk_score >= redact_at:
    redact
else:
    allow
```

This means a policy can block either because of accumulated score, a critical
finding, or a single rule that explicitly asks to block. A single high-severity
redaction finding does not block solely because its score equals a threshold.

## Redaction

Regex rules create match spans. When the resolved action is `redact`, or when a
redacting rule contributes to a report, matched spans are replaced with
`[REDACTED]`.

```text
input:
    Contact neel@example.com about the ticket.

output:
    Contact [REDACTED] about the ticket.
```

Function rules may not know exact character spans. They can still contribute
findings and affect the action, but span redaction depends on span data.

Use `redaction_strategy()` to change the replacement operator:

```{r}
scan_prompt(
  "Contact neel@example.com.",
  redaction = redaction_strategy("mask")
)

scan_prompt(
  "Contact neel@example.com.",
  redaction = redaction_strategy("hash")
)
```

Hash redaction is deterministic for the same matched text. It can support
linkage in audits, but it is not anonymization.

## Optional Scanners

Optional scanners run beside policy rules and return normal findings.

```{r}
scanners <- scanner_options(
  max_tokens = 500,
  blocked_topics = c("unreleased earnings"),
  allowed_url_hosts = c("example.com", "docs.example.com")
)

scan_prompt(
  "Email neel@example.com about unreleased earnings.",
  scanners = scanners
)
```

Default scanner options record invisible Unicode format characters and inspect
encoded payloads by decoding candidate URL-encoded and base64-like text, then
running policy rules over the decoded text. URL host policies, token limits,
language allowlists, and topic bans are enabled only when configured.

## Policy Controls

Scanner reports resolve to `allow`, `redact`, or `block`. `secure_chat()` then
uses policy controls to decide what to do with blocked surfaces.

```{r}
controlled <- policy(
  "enterprise_default",
  overrides = list(
    controls = policy_controls(
      on_prompt_block = "refuse",
      on_context_block = "drop",
      on_output_block = "escalate",
      refusal_message = "Please rephrase the request."
    )
  )
)
```

`on_context_block = "drop"` is the default because retrieved context is an
untrusted auxiliary input. Other context options are `keep_redacted`, `block`,
`refuse`, and `escalate`.

## Context Anomaly Scores

`scan_context()` adds RAG-specific findings before returning row-aligned
reports. It calculates:

- robust z-score of character length
- robust z-score of instruction-word density

Instruction density counts these words per 100 tokens:

```text
ignore, forget, override, instead, disregard
```

Rows above `anomaly_threshold`, default `2.5`, receive high-severity OWASP
LLM08 findings. If `source_col` is supplied and the policy has
`trusted_sources`, rows from outside that allowlist receive a medium-severity
OWASP LLM08 finding.

These RAG-specific findings are marked as synthetic. Their combined
contribution is capped at `0.3` per row before being added to normal rule
findings.

When context is passed through `secure_chat()`, included rows are assembled
with explicit `context row` labels, source labels when available, and separator
lines. This keeps retrieved text visually distinct from the user prompt and
preserves a row-level path back to audit data.

## Risk Summary

`secure_chat()` returns `risk_summary`, a named numeric vector keyed by OWASP
category.

```text
llm01 = prompt injection and instruction override findings
llm02 = sensitive information and secret findings
llm06 = excessive agency findings
llm08 = retrieved-context trust and anomaly findings
llm09 = misinformation and unsupported claim findings
llm10 = resource exhaustion and rate-guard findings
```

Each category is capped at `1.0`. The value is useful for dashboards and audit
summaries because it shows which risk category dominated a run.

## Rate Guard Semantics

`rate_guard()` now checks projected usage before counters are incremented.
`secure_chat()` reserves one request before the chat call. With
`strict = TRUE`, it also reserves an estimated prompt token count before the
call and records only the positive post-call delta after output scanning. If
the chat call or output scan fails, the pre-call reservation is rolled back.
`concurrent = TRUE` wraps each guard operation in a local file lock through the
optional `filelock` package.

## Extending a Policy

Start with the closest built-in policy, add rules, then inspect the resulting
inventory.

```{r}
guardrails <- policy()

guardrails <- add_rule(
  guardrails,
  id = "llm02.ticket_id",
  pattern = "\\bTICKET-[0-9]{6}\\b",
  owasp = "llm02",
  severity = "medium",
  action = "redact",
  description = "Internal support ticket identifier."
)

list_rules(guardrails)
```

When a policy becomes important to production, keep its custom rules in package
or application code, test them with representative prompts, and write audit
logs for real workflow runs.
