---
title: "Policy Design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Policy Design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This vignette explains how `llmshieldr` policies are assembled, what sources
the built-in presets draw from, and how the numeric scores become actions.

```{r}
library(llmshieldr)
```

## Design Goal

An LLM safety policy should be easy to inspect, easy to test, and easy to
extend. `llmshieldr` therefore uses explicit rule objects rather than a hidden
classifier as its default layer.

A policy is a list with these fields:

```text
shieldr_policy
    name             policy identifier stored in reports
    rules            list of shieldr_rule objects
    thresholds       redact_at and block_at numeric cutoffs
    rate_guard       optional shieldr_rate_guard environment
    trusted_sources  optional allowlist used by scan_context()
```

Each rule is similarly explicit:

```text
shieldr_rule
    id             stable rule identifier
    pattern        regex pattern, or NULL
    fn             R predicate function, or NULL
    owasp          OWASP LLM category
    severity       low, medium, high, or critical
    action         allow, redact, or block
    description    human-readable explanation
```

Exactly one of `pattern` or `fn` must be supplied. This keeps each rule
deterministic and makes redaction spans possible for regex rules.

## Source Model

The built-in rules are sourced from a small number of security and governance
concerns that recur across LLM applications:

- OWASP GenAI / LLM Top 10:
  <https://genai.owasp.org/llm-top-10/>
- Prompt-injection patterns: direct override language, hidden instructions,
  role confusion, and system-prompt extraction attempts
- Sensitive information patterns: email, phone, SSN, account numbers, medical
  record identifiers, subject IDs, API keys, bearer tokens, AWS keys, and
  credential-bearing connection strings
- Agentic risk patterns: model claims that it sent, deleted, granted,
  executed, notified, traded, or otherwise acted outside the chat boundary
- Output handling patterns: unsafe shell or SQL snippets, code execution
  markers, system-prompt structure, and high-confidence medical or financial
  claims
- RAG-specific signals: untrusted source metadata, anomalous chunk length, and
  high density of instruction words inside retrieved text
- Resource-exhaustion controls: request, token, and cost windows managed by
  `rate_guard()`

These sources are intentionally conservative. They are meant to catch common
failure modes in R workflows before prompts, retrieved context, or model output
cross a trust boundary.

## Preset Construction

Every preset is assembled in `policy_preset()`.

```text
enterprise_default
    prompt injection rules
    PII and secret rules
    system prompt extraction rules
    excessive agency rules

pharma_gxp
    enterprise_default
    MRN and clinical subject identifiers
    diagnosis and treatment claim language
    code-safety checks
    stricter thresholds: redact_at = 0.3, block_at = 0.6

finance_strict
    enterprise_default
    account number rules
    financial advice and guaranteed-return language
    autonomous investment-action language
    rate_guard(max_tokens = 100000, cost_limit_usd = 5.00)

education_safe
    enterprise_default
    minor-related PII patterns
    academic-integrity bypass language

open_research
    injection rules
    secret rules
    higher thresholds: redact_at = 0.8, block_at = 0.95

custom
    no rules
    default thresholds

baseline
    alias for enterprise_default
```

The policy names reflect intended operating posture, not legal compliance.
For example, `pharma_gxp` adds useful GxP-style controls, but it does not by
itself make a workflow compliant with any regulation.

## Scoring

Each finding contributes a numeric severity weight.

| Severity | Contribution |
| --- | ---: |
| `low` | 0.1 |
| `medium` | 0.3 |
| `high` | 0.6 |
| `critical` | 1.0 |

The scanner sums contributions and caps the total at `1.0`.

```text
findings:
    medium email finding      0.3
    high secret finding       0.6

risk_score = min(0.3 + 0.6, 1.0) = 0.9
```

The score is deliberately coarse. It is not a probability. It is a deterministic
severity index that helps a policy decide whether to allow, redact, or block.

## Action Resolution

Thresholds define how much risk the policy tolerates.

```text
default thresholds:
    redact_at = 0.40
    block_at  = 0.75
```

Action resolution is conservative:

```text
if any finding is critical:
    block
else if any finding's rule action is block:
    block
else if risk_score >= block_at:
    block
else if any finding's rule action is redact:
    redact
else if risk_score >= redact_at:
    redact
else:
    allow
```

This means a policy can block either because of accumulated score or because a
single high-risk rule explicitly asks to block.

## Redaction

Regex rules create match spans. When the resolved action is `redact`, or when a
redacting rule contributes to a report, matched spans are replaced with
`[REDACTED]`.

```text
input:
    Contact jane@example.com about the ticket.

output:
    Contact [REDACTED] about the ticket.
```

Function rules may not know exact character spans. They can still contribute
findings and affect the action, but span redaction depends on span data.

## Context Anomaly Scores

`scan_context()` adds RAG-specific findings before returning row-aligned
reports. It calculates:

- robust z-score of character length
- robust z-score of instruction-word density

Instruction density counts these words per 100 tokens:

```text
ignore, forget, override, instead, disregard
```

Rows above `anomaly_threshold`, default `2.5`, receive high-severity OWASP
LLM08 findings. If `source_col` is supplied and the policy has
`trusted_sources`, rows from outside that allowlist receive a medium-severity
OWASP LLM08 finding.

## Risk Summary

`secure_chat()` returns `risk_summary`, a named numeric vector keyed by OWASP
category.

```text
llm01 = prompt injection and instruction override findings
llm02 = sensitive information and secret findings
llm06 = excessive agency findings
llm08 = retrieved-context trust and anomaly findings
llm09 = misinformation and unsupported claim findings
llm10 = resource exhaustion and rate-guard findings
```

Each category is capped at `1.0`. The value is useful for dashboards and audit
summaries because it shows which risk category dominated a run.

## Extending a Preset

Start with the closest preset, add rules, then inspect the resulting inventory.

```{r}
policy <- policy_preset("enterprise_default")

policy <- add_rule(
  policy,
  id = "llm02.ticket_id",
  pattern = "\\bTICKET-[0-9]{6}\\b",
  owasp = "llm02",
  severity = "medium",
  action = "redact",
  description = "Internal support ticket identifier."
)

list_rules(policy)
```

When a policy becomes important to production, keep its custom rules in package
or application code, test them with representative prompts, and write audit
logs for real workflow runs.
