---
title: "Gemini and Ollama Workflows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gemini and Ollama Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)

.llmshieldr_is_check <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))
.llmshieldr_is_github_actions <- identical(
  tolower(Sys.getenv("GITHUB_ACTIONS", unset = "")),
  "true"
)
.llmshieldr_has_ellmer <- requireNamespace("ellmer", quietly = TRUE)
.llmshieldr_has_gemini <-
  !.llmshieldr_is_check &&
  !.llmshieldr_is_github_actions &&
  .llmshieldr_has_ellmer &&
  (
    nzchar(Sys.getenv("GEMINI_API_KEY")) ||
      nzchar(Sys.getenv("GOOGLE_API_KEY"))
  )

.llmshieldr_ollama_models <- NULL
if (
  !.llmshieldr_is_check &&
    !.llmshieldr_is_github_actions &&
    .llmshieldr_has_ellmer
) {
  .llmshieldr_ollama_models <- tryCatch(
    suppressWarnings(suppressMessages(ellmer::models_ollama())),
    error = function(error) NULL
  )
}
.llmshieldr_has_ollama <- !is.null(.llmshieldr_ollama_models) &&
  "id" %in% names(.llmshieldr_ollama_models) &&
  length(.llmshieldr_ollama_models$id) > 0L &&
  !is.na(.llmshieldr_ollama_models$id[[1]]) &&
  nzchar(.llmshieldr_ollama_models$id[[1]])

.llmshieldr_ollama_model <- trimws(
  Sys.getenv("llmshieldr_ollama_model", unset = "")
)
if (!nzchar(.llmshieldr_ollama_model) && .llmshieldr_has_ollama) {
  .llmshieldr_ollama_model <-
    as.character(.llmshieldr_ollama_models$id[[1]])
}
if (!nzchar(.llmshieldr_ollama_model)) {
  .llmshieldr_ollama_model <- NULL
}

.llmshieldr_run_example <- function(provider, code) {
  old_options <- options(
    cli.progress_show_after = Inf,
    cli.unicode = FALSE
  )
  on.exit(options(old_options), add = TRUE)

  tryCatch(
    force(code),
    error = function(error) {
      first_line <- trimws(
        strsplit(conditionMessage(error), "\n", fixed = TRUE)[[1]][[1]]
      )
      message(provider, " example was skipped: ", first_line)
      invisible(NULL)
    }
  )
}
```

`secure_chat()` is the common path for Gemini, Ollama, and other providers
supported by `ellmer`. This guide intentionally runs no more than one request
per provider.

Live calls are disabled during package checks and whenever
`GITHUB_ACTIONS=true`. A missing provider, unavailable model, quota error, or
temporary service failure is reported without failing the documentation build.

```{r provider-availability, echo = FALSE, eval = TRUE}
data.frame(
  provider = c("Gemini", "Ollama"),
  live_example = c(
    .llmshieldr_has_gemini,
    .llmshieldr_has_ollama
  )
)
```

## Install ellmer

```{r install-ellmer, eval = FALSE}
install.packages("ellmer")
```

```{r load-live-providers, include = FALSE, eval = .llmshieldr_has_gemini || .llmshieldr_has_ollama}
library(llmshieldr)
```

## Gemini

Add one API key to the user-level `~/.Renviron` file:

```text
GEMINI_API_KEY=replace-with-your-real-key
```

`GOOGLE_API_KEY` is also accepted. Use the exact filename `.Renviron`, restart
R after saving it, and never commit the file or print the key.

The following chunk runs locally only when `ellmer` and a key are available:

```{r gemini-example, eval = .llmshieldr_has_gemini}
gemini_result <- .llmshieldr_run_example(
  "Gemini",
  secure_chat(
    prompt = "Explain in one sentence why LLM output should be checked.",
    provider = "gemini",
    model = "gemini-3.8-flash",
    checks = "rules",
    show_stats = TRUE
  )
)

if (!is.null(gemini_result)) {
  data.frame(
    action = gemini_result$action,
    output = gemini_result$output
  )
}
```

Google controls model availability, quotas, pricing, regions, and data-use
terms. Confirm the current terms before sending private or regulated data.

## Ollama

Start Ollama and pull a model outside R. For example:

```text
ollama serve
ollama pull llama3.2
```

Choose the model in `~/.Renviron`:

```text
llmshieldr_ollama_model=llama3.2
```

Restart R after changing the file. When this variable is absent, the vignette
uses the first model returned by `ellmer::models_ollama()`.

The following chunk runs locally only when Ollama is reachable and has a model:

```{r ollama-example, eval = .llmshieldr_has_ollama}
ollama_result <- .llmshieldr_run_example(
  "Ollama",
  secure_chat(
    prompt = "Explain in one sentence why LLM output should be checked.",
    provider = "ollama",
    model = .llmshieldr_ollama_model,
    checks = "rules",
    show_stats = TRUE
  )
)

if (!is.null(ollama_result)) {
  data.frame(
    action = ollama_result$action,
    output = ollama_result$output
  )
}
```

## Recommended Next Steps

After the basic request works, try one change at a time:

| Goal | Recommendation |
|---|---|
| Add semantic review | Set `checks = "both"` and provide `reviewer_model`. Start with a model that reliably returns JSON. |
| Guard retrieved documents | Add `context` and a `context_policy()` before trying a full RAG workflow. |
| Constrain responses | Add an `output_contract()` for the required format and size. |
| Inspect usage | Set `show_tokens = TRUE` and keep `show_stats = TRUE`. |
| Reuse an existing client | Pass an existing ellmer object through `chat` instead of `provider`. |
| Guard tool use | Register only required tools and set `allowed_tools` explicitly. |
| Evaluate a policy | Create benign and adversarial cases, then call `evaluate_security_cases()`. |

The RAG, finance, pharma, policy design, and custom rules vignettes show these
features without making additional provider calls during rendering.

## Provider-Specific Wrappers

`shield_gemini()` and `shield_ollama()` remain available for compatibility and
issue a deprecation warning. Use `secure_chat(provider = ...)` in new code.

## Troubleshooting

- **Gemini key not found:** confirm the variable is in `~/.Renviron`, then
  restart R.
- **Gemini quota or model error:** check the active Google project, current
  quota, and model availability.
- **Ollama unavailable:** start the service and confirm `ollama list` returns
  the configured model.
- **Semantic reviewer returns malformed JSON:** try a stronger local model or
  define the required failure behavior with `policy_controls()`.
